---
title: "p8105_hw1_NC2710"
author: "Nicole Comfort"
date: "9/18/2018"
output: github_document
---

```{r setup, include=FALSE}
remove(list=ls()) # start with clear environment
get(wd) # show that I'm in a reasonable working directory
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

## Problem 0.1

This “problem” focuses on the use of R Markdown to write reproducible reports, GitHub for version control, and R Projects to organize your work.

To address Problem 0.1, I've:
* created a public GitHub repo titled p8105_hw1_NC2710 + local R Project
* create a single .Rmd file named p8105_hw1_NC2710.Rmd that renders to github_document

## Problem 0.2

To address Problem 0.2, I will make sure to have:
* meaningful variable / object names
* readable code (one command per line; adequate whitespace and indentation; etc)
* clearly-written text to explain code and results
* a lack of superfluous code (e.g. no unused variables are defined)

## Problem 1 

This problem focuses on variable types and coercion.

The code chunk below creates a data frame comprised of: 
* A random sample of size 10 from a uniform[0, 5] distribution
* A logical vector indicating whether elements of the sample are greater than 2
* A (length-10) character vector
* A (length-10) factor vector

```{r Problem 1 data frame}

set.seed(1) # ensures reproducibility when using the random number generator, rnorm

prob1_df = tibble( # create data frame
  prob1_rand_sample = rnorm(10, mean = 0, sd = 5), # create a random sample of size 10 from a uniform[0,5] distribution
  prob1_vec_logical = prob1_rand_sample > 2, # create a logical vector which evalues prob1_rand_sample
  prob1_vec_char = c("This", "is", "an", "example", "character", "vector", "with", "length", "equals", "ten"), # create an example character vector of length 10
  prob1_vec_factor = factor(c("red", "orange", "yellow", "green", "blue", "cyan", "purple", "indigo", "violet", "gray")) # create a factor vector of length 10 (i.e. with 10 levels) 
)

```

*Try to take the mean of each variable in your dataframe. What works and what doesn’t? Why?* 

The below code chunk attempts to take the mean of each variable in the dataframe. However, the function is only executed for the variables, prob1_rand_sample and prob1_vec_logical, which are of the types numeric and logical, respectively. 

I am surprised to see that R can take the mean of a logical variable. It appears that R treats "TRUE's" as "1"s and "FALSE's" as "0's" to compute the mean. 

```{r Problem 1 means}

mean(prob1_df$prob1_rand_sample) # mean works
mean(prob1_df$prob1_vec_logical) # mean works
mean(prob1_df$prob1_vec_char) # mean does not work, returns NA because the argument is not numeric or logical
mean(prob1_df$prob1_vec_factor) # mean does not work, returns NA because the argument is not numeric or logical

```

*In some cases, you can explicitly convert variables from one type to another. Write a code chunk that applies the as.numeric function to the logical, character, and factor variables (please show this chunk but not the output). What happens? In a second code chunk, convert your character variable from character to factor to numeric; similarly, convert your factor variable from factor to character to numeric. What happens?* 


